DecipherPD Mouse Enriched Environment Hippocampus Differential Methylation
================================

##### Analyst: Samantha Schaffner
##### Date: Jan 16, 2023

In this script I will compare DNAm and DNAhm levels at CpGs with differential hydroxymethylation. Typically these should be inversely correlated; however, loss of hmC is possible without gain of mC, if exisiting hmC is oxidized to C. I'll be looking more for a loss of mC associated with gain of hmC.

## Load in data and set working directories

```{r load data, eval=F}
.libPaths(c(.libPaths(),"/mnt/scratch/KoborLab/R_Libs/4.2"))
library(BiSeq)
library(GenomicRanges)
library(ggplot2)
library(reshape2)
library(dplyr)
library(gplots)
library(RColorBrewer)
library(gtools)
library(DescTools)
library(normtest)
library(ggpubr)
library(gridExtra)
library(bacon)
library(NCmisc)
library(limma)
library(car)
library(gtools)
library(RColorBrewer)
library(pvclust)
source("/mnt/scratch/KoborLab/shared_coding_resource/qqplot_NG.R")
```

## Reading in differentially hydroxymethylated CpGs list

```{r volcano 1, eval=F}
results_G <- read.csv("/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/082022/5-methdiff_DNAm_DNAhm/hydroxy/WTSE_TGSE/linear_model/auto/CpG/lm_WTSE_TGSE.csv")

#CpGs with increased hmC in TG mice
hmC_up <- results_G[results_G$DB>0 & results_G$threshold==TRUE,"site"] #61 CpGs

#CpGs with decreased hmC in TG mice
hmC_down <- results_G[results_G$DB<0 & results_G$threshold==TRUE,"site"] #1522 CpGs

load("/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/082022/5-methdiff_DNAm_DNAhm/preprocessing/oxbisulfite/CpG/all_EE_mC_betas_CpG_10X.RData")
dim(all_EE)
#[1] 914215     16
colnames(all_EE)
all_EE <- all_EE[,-c(8,12,13)] #removing outliers

####### CpGs with increased hmC in TG mice (expecting decreased mC in TG mice)
#DNAm
nrow(betas_mC <- all_EE[rownames(all_EE) %in% hmC_up,]) #61

#DNAhm
load("/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/082022/5-methdiff_DNAm_DNAhm/preprocessing/hydroxy/CpG/all_EE_hmC_betas_CpG_10X_auto_var_noout.RData")
nrow(betas_hmC <- betas_variable[rownames(betas_variable) %in% hmC_up,]) #61
all(rownames(betas_mC)==rownames(betas_hmC)) #TRUE

colnames(betas_mC) <- paste(colnames(betas_mC), "mC", sep=".")
colnames(betas_hmC) <- paste(colnames(betas_hmC), "hmC", sep=".")
betas_combined <- cbind(betas_mC, betas_hmC)

betas_combined <- as.matrix(betas_combined)
Spect <- brewer.pal("Spectral", n=11)
Greys <- brewer.pal("Greys", n=9)

#clustering and significance testing
clust <- pvclust(betas_sub)
plot(clust)

heatmap.2(betas_combined, trace="none", col=Greys, Colv=NA, symbreaks=F, symkey=F)
#WTSE has the lowest hmC. The TG mice have a range of different mC values. Around half have loss of mC in TG mice.

####### CpGs with decreased hmC in TG mice 
#DNAm
nrow(betas_mC <- all_EE[rownames(all_EE) %in% hmC_down,]) #1522

#DNAhm
nrow(betas_hmC <- betas_variable[rownames(betas_variable) %in% hmC_down,]) #1522
all(rownames(betas_mC)==rownames(betas_hmC)) #TRUE

colnames(betas_mC) <- paste(colnames(betas_mC), "mC", sep=".")
colnames(betas_hmC) <- paste(colnames(betas_hmC), "hmC", sep=".")
betas_combined <- cbind(betas_mC, betas_hmC)

betas_combined <- as.matrix(betas_combined)

heatmap.2(betas_combined, trace="none", col=Greys, Colv=NA, symbreaks=F, symkey=F)

####### Check significance of DHM CpGs in differential methylation analysis
# nominal level since not much differential methylation was seen
results_mC <- read.csv("/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/082022/5-methdiff_DNAm_DNAhm/methylation/WTSE_TGSE/linear_model/auto/CpG/lm_WTSE_TGSE.csv")
nrow(mC_change <- results_mC[results_mC$Pr...t..<=0.01 & abs(results_mC$DB)>=0.1,]) #2674
length(mC_hmC <- mC_change$site[mC_change$site %in% hmC_down|mC_change$site %in% hmC_up]) #80
summary(results_mC[results_mC$site %in% mC_hmC,c("Pr...t..","adjP")])
#    Pr...t..              adjP       
# Min.   :1.171e-05   Min.   :0.3129  
# 1st Qu.:1.834e-03   1st Qu.:0.8724  
# Median :3.976e-03   Median :0.9275  
# Mean   :4.426e-03   Mean   :0.9004  
# 3rd Qu.:7.058e-03   3rd Qu.:0.9730  
# Max.   :9.954e-03   Max.   :1.0000 

#though nominally significant, these mC changes really don't come close to passing MTC, with adjusted p-values close to 0.9 on average

summary(abs(results_mC[results_mC$site %in% mC_hmC,"DB"]))
#   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 0.1141  0.1768  0.2290  0.2241  0.2543  0.3862 

summary(abs(results_G[results_G$site %in% mC_hmC,"DB"]))
#   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 0.1280  0.2170  0.2697  0.2675  0.3059  0.4878 

nrow(betas_mC <- all_EE[rownames(all_EE) %in% mC_hmC,]) #80
nrow(betas_hmC <- betas_variable[rownames(betas_variable) %in% mC_hmC,]) #80
all(rownames(betas_mC)==rownames(betas_hmC)) #TRUE

colnames(betas_mC) <- paste(colnames(betas_mC), "mC", sep=".")
colnames(betas_hmC) <- paste(colnames(betas_hmC), "hmC", sep=".")
betas_combined <- cbind(betas_mC, betas_hmC)

betas_combined <- as.matrix(betas_combined)

heatmap.2(betas_combined, trace="none", col=Spect, Colv=NA, symbreaks=F, symkey=F)
```

