# Creating background annotation files for genomic and CpG island contexts
### Author: Samantha Schaffner
### Date: Dec 12, 2022

Most cytosine sites are annotated to multiple features. In this script, I create a subsetted annotation where each site is annotated to the feature corresponding to its longest transcript, in order to capture the widest possible range of gene regulation.

These background annotation files can be used for downstream analysis such as genomic feature enrichment and gene ontology enrichment.

Note: this code should be run after completing "transcript_length.Rmd."

##Load in your fully annotated object and longest transcript annotation
The objects "results_anno.RData" and "longest_transcript.RData" were generated in the script "transcript_length.Rmd."
```{r load objects, eval=F}
.libPaths(c(.libPaths(), "/mnt/scratch/KoborLab/R_Libs/4.2", "/mnt/scratch/KoborLab/Personal_Folders/sschaffner/Rstudio/R/x86_64-pc-linux-gnu-library/4.0/"))
library(lifecycle, lib.loc="/mnt/scratch/KoborLab/R_Libs/4.2")
library(scales, lib.loc="/mnt/scratch/KoborLab/R_Libs/4.2")
library(vctrs, lib.loc="/mnt/scratch/KoborLab/R_Libs/4.2")

library(annotatr)
library(AnnotationHub)
library(BiSeq)
#library(biomaRt)

load("/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/082022/5-methdiff_DNAm_DNAhm/preprocessing/hydroxy/CpG/hmC_anno_CpG.RData")
load("/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/082022/5-methdiff_DNAm_DNAhm/preprocessing/hydroxy/CpG/longest_transcript_hmC_CpG.RData")

#create a unique transcript-cpg ID for matching
hmC_anno$annot.tx_id <- sapply(1:nrow(hmC_anno), function(x) { unlist(strsplit(hmC_anno$annot.tx_id[x], split="\\."))[1]})
hmC_anno$tx_site <- paste(hmC_anno$site, hmC_anno$annot.tx_id, sep=".")
longest_transcript[longest_transcript=="none"] <- NA
longest_transcript <- longest_transcript[complete.cases(names(longest_transcript))]

#How many sites have a transcript?
length(longest_transcript[complete.cases(longest_transcript)]) #13,674

longest_tx_site <- paste(names(longest_transcript), longest_transcript, sep=".")
nrow(hmC_anno_sub <- hmC_anno[hmC_anno$tx_site %in% longest_tx_site,]) #30,898
length(unique(hmC_anno_sub$site)) #19,308
length(unique(hmC_anno$site)) #19,308 - all sites have at least one annotation

#Even after subsetting the annotation to the longest transcript, there are multiple features annotated per site.
```

#For those sites which are still annotated to multiple features, collapse the annotation to one feature per site

Even though we already subset to longest transcript, some of the annotation categories from the annotatr package overlap with each other by definition. For example, a first exon could be labelled as both "mm10_genes_exons" and "mm10_genes_firstexons." See http://bioconductor.org/packages/release/bioc/vignettes/annotatr/inst/doc/annotatr-vignette.html#genic-annotations for a visual representation of where the annotar categories lie within a gene.

In this step, I prioritize the annotation categories from most specific to most general. For example, if a feature is located in an exon-intron boundary, I'll assign that as the most specific feature for that sites.
```{r specific anno, eval=F}
#subset further to get a list of unique cytosine sites in the annotation
results_anno_multi_genes_sub <- hmC_anno_sub[,c(1,2,3,12,13,14,15,16)]
#remove any duplicate rows
results_anno_multi_genes_sub <- distinct(results_anno_multi_genes_sub)
length(unique_cpgs <- unique(results_anno_multi_genes_sub$site)) #19,308

#Within each longest transcript annotation, determine what features would over-ride each other...
#Focusing on gene rather than CpG island features first
unique(results_anno_multi_genes_sub$annot.type)
# [1] "mm10_genes_introns"              "mm10_genes_exonintronboundaries" "mm10_genes_intronexonboundaries" "mm10_genes_intergenic"          
# [5] "mm10_cpg_inter"                  "mm10_genes_exons"                "mm10_genes_3UTRs"                "mm10_genes_firstexons"          
# [9] "mm10_genes_promoters"            "mm10_lncrna_gencode"             "mm10_genes_1to5kb"               "mm10_genes_5UTRs"               
#[13] "mm10_cpg_shelves"                "mm10_cpg_shores"                 "mm10_cpg_islands"       

#Go from the most specific annotation to the most general annotation.
specific_gene <- rep(NA, length(unique_cpgs))
specific_tx <- rep(NA, length(unique_cpgs))
specific_anno <- rep(NA, length(unique_cpgs))
  
for (x in 1:length(unique_cpgs)){
  print(x)
  cpg_set <- results_anno_multi_genes_sub[results_anno_multi_genes_sub$site==unique_cpgs[x],]
  if (nrow(cpg_set)>1){
    if ("mm10_lncrna_gencode" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_lncrna_gencode","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_lncrna_gencode","annot.symbol"])
      specific_anno[x] <- "mm10_lncrna_gencode"
    }
    if ("mm10_genes_exonintronboundaries" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_exonintronboundaries","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_exonintronboundaries","annot.symbol"])
      specific_anno[x] <- "mm10_genes_exonintronboundaries"
    }
    else if ("mm10_genes_intronexonboundaries" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_intronexonboundaries","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_intronexonboundaries","annot.symbol"])
      specific_anno[x] <- "mm10_genes_intronexonboundaries"
    }
    else if ("mm10_genes_promoters" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_promoters","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_promoters","annot.symbol"])
      specific_anno[x] <- "mm10_genes_promoters"
    }
    else if ("mm10_genes_firstexon" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_firstexon","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_firstexon","annot.symbol"])
      specific_anno[x] <- "mm10_genes_firstexon"
    }
    else if ("mm10_genes_5UTRs" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_5UTRs","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_5UTRs","annot.symbol"])
      specific_anno[x] <- "mm10_genes_5UTRs"
    }
    else if ("mm10_genes_3UTRs" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_3UTRs","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_3UTRs","annot.symbol"])
      specific_anno[x] <- "mm10_genes_3UTRs"
    }
    else if ("mm10_genes_1to5kb" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_1to5kb","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_1to5kb","annot.symbol"])
      specific_anno[x] <- "mm10_genes_1to5kb"
    }
    else if ("mm10_genes_exons" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_exons","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_exons","annot.symbol"])
      specific_anno[x] <- "mm10_genes_exons"
    }
    else if ("mm10_genes_introns" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_introns","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_introns","annot.symbol"])
      specific_anno[x] <- "mm10_genes_introns"
    }
    else if ("mm10_genes_cds" %in% cpg_set$annot.type){
      specific_tx[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_cds","annot.tx_id"])
      specific_gene[x] <- unique(cpg_set[cpg_set$annot.type=="mm10_genes_cds","annot.symbol"])
      specific_anno[x] <- "mm10_genes_cds"
    }
  }
  else if (nrow(cpg_set)==1){ 
    specific_tx[x] <- cpg_set$annot.tx_id[x]
    specific_gene[x] <- cpg_set$annot.symbol[x]
    specific_anno[x] <- cpg_set$annot.type }
  
}

nrow(cpg_anno <- data.frame(site=unique_cpgs, gene=specific_gene, tx=specific_tx, genomic_anno=specific_anno)) #19,308
unique(cpg_anno$genomic_anno) #some CpG island annotations are still in there; these would be intergenic sites (above loop would catch any other gene-associated annotations present)
cpg_anno$genomic_anno[cpg_anno$genomic_anno %in% c("mm10_cpg_islands", "mm10_cpg_shores","mm10_cpg_shelves","mm10_cpg_inter")] <- NA
cpg_anno$genomic_anno <- as.character(cpg_anno$genomic_anno)
cpg_anno$genomic_anno[is.na(cpg_anno$genomic_anno)] <- "mm10_genes_intergenic"
```

##Add CpG island context annotation
```{r cgi context, eval=F}
#Now, annotate CpG island context.
#cpg islands, etc. annotation is only in the results without genes
nrow(results_nogenes <- hmC_anno[hmC_anno$annot.type %in% c("mm10_cpg_inter", "mm10_cpg_islands", "mm10_cpg_shelves", "mm10_cpg_shores"),]) #19,308
cpg_anno_cgi <- results_nogenes

cpg_anno_cgi <- cpg_anno_cgi[match(cpg_anno$site, cpg_anno_cgi$site),]
all.equal(cpg_anno_cgi$site, as.character(cpg_anno$site)) #TRUE
cpg_anno$cgi_anno <- cpg_anno_cgi$annot.type

write.table(cpg_anno, file="/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/082022/5-methdiff_DNAm_DNAhm/preprocessing/hydroxy/CpG/cpg_anno_hmC.txt", sep="\t")
```

##Enhancer and promoter annotations from ChIP-seq
Added Oct 6, 2021. See "enhancers_ZW.R" and "promoters_ZW.R" in KoborLab/kobor_space/sschaffner/DecipherPD_RRBS/SNCA_OE/Hippocampus/12m_EE/ChIPseq/Zinah_Wassouf_data/consensus_peaks.
```{r enh prom, eval=F}
enactive <- read.delim("/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/7-ChIPseq/Zinah_Wassouf_data/consensus_peaks/enhancers_active.txt")
enpoised <- read.delim("/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/7-ChIPseq/Zinah_Wassouf_data/consensus_peaks/enhancers_poised.txt")
promoters <- read.delim("/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/7-ChIPseq/Zinah_Wassouf_data/consensus_peaks/promoters.txt")

unique(cpg_anno$genomic_anno)
# [1] "mm10_genes_introns"              "mm10_genes_exonintronboundaries" "mm10_genes_intergenic"           "mm10_genes_exons"               
# [5] "mm10_genes_intronexonboundaries" "mm10_genes_promoters"            "mm10_genes_3UTRs"                "mm10_genes_1to5kb"              
# [9] "mm10_genes_5UTRs"                "mm10_lncrna_gencode"          

library(GenomicRanges)
cpg_anno$site <- as.character(cpg_anno$site)
cpg_ranges <- GRanges(seqnames=sapply(cpg_anno$site, function(x) unlist(strsplit(x, split=": "))[1]), ranges=IRanges(start=as.numeric(sapply(cpg_anno$site, function(x) unlist(strsplit(x, split=": "))[2])), end=as.numeric(sapply(cpg_anno$site, function(x) unlist(strsplit(x, split=": "))[2]))), mcols=cpg_anno$site)

enactive_ranges <- GRanges(seqnames=enactive$seqnames, ranges=IRanges(start=enactive$start, end=enactive$end))
enpoised_ranges <- GRanges(seqnames=enpoised$seqnames, ranges=IRanges(start=enpoised$start, end=enpoised$end))
promoters_ranges <- GRanges(seqnames=promoters$seqnames, ranges=IRanges(start=promoters$start, end=promoters$end))

cpg_enactive <- as.data.frame(subsetByOverlaps(cpg_ranges, enactive_ranges))
cpg_enpoised <- as.data.frame(subsetByOverlaps(cpg_ranges, enpoised_ranges))
cpg_promoters <- as.data.frame(subsetByOverlaps(cpg_ranges, promoters_ranges))

#checking how the old annotations map onto the new ones
cpg_anno$genomic_anno <- as.factor(cpg_anno$genomic_anno)

#active enhancers
summary(cpg_anno[cpg_anno$site %in% cpg_enactive$mcols,"genomic_anno"])
#              mm10_genes_1to5kb                mm10_genes_3UTRs                mm10_genes_5UTRs mm10_genes_exonintronboundaries 
#                             78                              44                               2                             331 
#               mm10_genes_exons           mm10_genes_intergenic mm10_genes_intronexonboundaries              mm10_genes_introns 
#                            106                             331                             107                            1443 
#           mm10_genes_promoters             mm10_lncrna_gencode 
#                             33                               1 

#poised enhancers
summary(cpg_anno[cpg_anno$site %in% cpg_enpoised$mcols,"genomic_anno"])
#              mm10_genes_1to5kb                mm10_genes_3UTRs                mm10_genes_5UTRs mm10_genes_exonintronboundaries 
#                              4                               7                               3                              65 
#               mm10_genes_exons           mm10_genes_intergenic mm10_genes_intronexonboundaries              mm10_genes_introns 
#                             13                             115                               5                             237 
#           mm10_genes_promoters             mm10_lncrna_gencode 
#                              2                               0 

#promoters
summary(cpg_anno[cpg_anno$site %in% cpg_promoters$mcols,"genomic_anno"])
#              mm10_genes_1to5kb                mm10_genes_3UTRs                mm10_genes_5UTRs mm10_genes_exonintronboundaries 
#                            218                             113                              11                             503 
#               mm10_genes_exons           mm10_genes_intergenic mm10_genes_intronexonboundaries              mm10_genes_introns 
#                            227                             172                             159                             710 
#           mm10_genes_promoters             mm10_lncrna_gencode 
#                            124                              11 


#using ChIP-seq annotations as the most "specific"
str(cpg_anno$genomic_anno) #should be character
cpg_anno$genomic_anno <- as.character(cpg_anno$genomic_anno)
cpg_anno[cpg_anno$site %in% cpg_enactive$mcols,"genomic_anno"] <- "enhancers_active"
cpg_anno[cpg_anno$site %in% cpg_enpoised$mcols,"genomic_anno"] <- "enhancers_poised"
cpg_anno[cpg_anno$site %in% cpg_promoters$mcols,"genomic_anno"] <- "promoters_active"

unique(cpg_anno$genomic_anno)
# [1] "enhancers_poised"                "mm10_genes_introns"              "mm10_genes_exonintronboundaries" "mm10_genes_intergenic"          
# [5] "enhancers_active"                "mm10_genes_intronexonboundaries" "promoters_active"                "mm10_genes_promoters"           
# [9] "mm10_genes_3UTRs"                "mm10_genes_1to5kb"               "mm10_genes_exons"                "mm10_lncrna_gencode"            
#[13] "mm10_genes_5UTRs"        

write.table(cpg_anno, file="/mnt/scratch/KoborLab/DecipherPD_mouse/ee_asyn/hipp/sschaffner/082022/5-methdiff_DNAm_DNAhm/preprocessing/hydroxy/CpG/cpg_anno_hmC_ChIP.txt", sep="\t")
```

