In this script I will look at the average DNA methylation across lambda phage spike-in DNA to determine bisulfite conversion efficiency. Phage DNA should all be unmethylated, and high quality samples should show an average methylation of 2% or less.

Checking outliers first: WT_12m_EE_4, TG_12m_EE_2, TG_12m_std_1

WT 12m std
```{r}
WT_12m_std_1 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0869_oxBS_pool1/ACAGTG_methratio_lambda_only.txt")
mean(WT_12m_std_1$ratio) #0.006708479

WT_12m_std_2 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0869_oxBS_pool1/ACTTGA_methratio_lambda_only.txt")
mean(WT_12m_std_2$ratio) #0.007476594

WT_12m_std_3 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0869_oxBS_pool1/ATCACG_methratio_lambda_only.txt")
mean(WT_12m_std_3$ratio) #0.01060752

WT_12m_std_4 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0869_oxBS_pool1/CAGATC_methratio_lambda_only.txt")
mean(WT_12m_std_4$ratio) #0.009279661
```
WT 12m EE
```{r}
WT_12m_EE_1 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0869_oxBS_pool1/CGATGT_methratio_lambda_only.txt")
mean(WT_12m_EE_1$ratio) #0.007564729

WT_12m_EE_2 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0869_oxBS_pool1/GCCAAT_methratio_lambda_only.txt")
mean(WT_12m_EE_2$ratio) #0.008773356

WT_12m_EE_3 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0869_oxBS_pool1/TGACCA_methratio_lambda_only.txt")
mean(WT_12m_EE_3$ratio) #0.0079125

WT_12m_EE_4 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0869_oxBS_pool1/TTAGGC_methratio_lambda_only.txt")
mean(WT_12m_EE_4$ratio) #0.008823061
```
TG 12m std
```{r}
TG_12m_std_1 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0870_oxBS_pool2/ACAGTG_methratio_lambda_only.txt")
mean(TG_12m_std_1$ratio) #0.007886425

TG_12m_std_2 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0870_oxBS_pool2/ACTTGA_methratio_lambda_only.txt")
mean(TG_12m_std_2$ratio) #0.008868263

TG_12m_std_3 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0870_oxBS_pool2/ATCACG_methratio_lambda_only.txt")
mean(TG_12m_std_3$ratio) #0.008858986

TG_12m_std_4 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0870_oxBS_pool2/CAGATC_methratio_lambda_only.txt")
mean(TG_12m_std_4$ratio) #0.008204137
```
TG 12m EE
```{r}
TG_12m_EE_1 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0870_oxBS_pool2/CGATGT_methratio_lambda_only.txt")
mean(TG_12m_EE_1$ratio) #0.006677606

TG_12m_EE_2 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0870_oxBS_pool2/GCCAAT_methratio_lambda_only.txt")
mean(TG_12m_EE_2$ratio) #0.00653367

TG_12m_EE_3 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0870_oxBS_pool2/TGACCA_methratio_lambda_only.txt")
mean(TG_12m_EE_3$ratio) #0.007410984

TG_12m_EE_4 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0870_oxBS_pool2/TTAGGC_methratio_lambda_only.txt")
mean(TG_12m_EE_4$ratio) #0.006608163
```
WT 6m std
```{r}
WT_6m_std_1 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0871_oxBS_pool3/ACAGTG_methratio_lambda_only.txt")
mean(WT_6m_std_1$ratio) #0.008069987

WT_6m_std_2 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0871_oxBS_pool3/ACTTGA_methratio_lambda_only.txt")
mean(WT_6m_std_2$ratio) #0.007918704

WT_6m_std_3 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0871_oxBS_pool3/ATCACG_methratio_lambda_only.txt")
mean(WT_6m_std_3$ratio) #0.007386679

WT_6m_std_4 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0871_oxBS_pool3/CAGATC_methratio_lambda_only.txt")
mean(WT_6m_std_4$ratio) #0.006733965
```
WT 6m stress
```{r}
WT_6m_stress_1 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0871_oxBS_pool3/CGATGT_methratio_lambda_only.txt")
mean(WT_6m_stress_1$ratio) #0.00543447

WT_6m_stress_2 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0871_oxBS_pool3/GCCAAT_methratio_lambda_only.txt")
mean(WT_6m_stress_2$ratio) #0.008499023

WT_6m_stress_3 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0871_oxBS_pool3/TGACCA_methratio_lambda_only.txt")
mean(WT_6m_stress_3$ratio) #0.005296136

WT_6m_stress_4 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0871_oxBS_pool3/TTAGGC_methratio_lambda_only.txt")
mean(WT_6m_stress_4$ratio) #0.01070083
```

TG 6m std
```{r}
TG_6m_std_1 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0872_oxBS_pool4/ACAGTG_methratio_lambda_only.txt")
mean(TG_6m_std_1$ratio) #0.007937158

TG_6m_std_2 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0872_oxBS_pool4/ACTTGA_methratio_lambda_only.txt")
mean(TG_6m_std_2$ratio) #0.007868603

TG_6m_std_3 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0872_oxBS_pool4/ATCACG_methratio_lambda_only.txt")
mean(TG_6m_std_3$ratio) #0.01049263

TG_6m_std_4 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0872_oxBS_pool4/CAGATC_methratio_lambda_only.txt")
mean(TG_6m_std_4$ratio) #0.009302289
```
TG 6m stress
```{r}
TG_6m_stress_1 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0872_oxBS_pool4/CGATGT_methratio_lambda_only.txt")
mean(TG_6m_stress_1$ratio) #0.00806334

TG_6m_stress_2 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0872_oxBS_pool4/GCCAAT_methratio_lambda_only.txt")
mean(TG_6m_stress_2$ratio) #0.008365525

TG_6m_stress_3 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0872_oxBS_pool4/TGACCA_methratio_lambda_only.txt")
mean(TG_6m_stress_3$ratio) #0.007946369

TG_6m_stress_4 <- read.delim("~/KoborLab/Sam_New_Test_Data/RRBS/methratio/PX0872_oxBS_pool4/TTAGGC_methratio_lambda_only.txt")
mean(TG_6m_stress_4$ratio) #0.008561485
```

All the oxBS samples had successful bisulfite conversion (average DNAm in lambda spike-in was no greater than 1%).